\stdpopsim: a community toolkit for population genetic simulation

### 2019-12-05 popSimManu.tex
shortened the title
KEL edits on abstract & intro

\title{\stdpopsim: a community-supported toolkit for population genetic simulation}

@@

\abstract{The explosion in population genomic data demands ever more complex
modes of analysis, and increasingly these analyses depend on sophisticated simulations.
Recent advances in population genetic simulation have made it possible to
simulate large and complex models, but describing such models for a particular
simulation engine remains a difficult and error-prone task.
Computational genetics researchers currently re-implement simulation models
independently, leading to duplication of effort and reductions in reproducibility.
Population genetics, as a field, also lacks rigorous standard benchmarks by which new tools for inference might
be measured. Here we describe a new resource, stdpopsim, that attempts to rectify this situation.
Stdpopsim is a community-driven, open source project, which provides a standard
catalog of published simulation models from a wide range of organisms,
and supports multiple simulation engine backends.
% PopSim assures correctness in our implementations through a mature quality control pipeline that is community-based.
Here we describe the stdpopsim resource to encourange an even broader
community of developers to participate and to demonstrate some examples of how
it can be used to systematically compare demographic inference methods.}
\date{\today}

@@

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, the extent of positive selection in genome sequence data
(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}).
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to
introduce methods which are validated by a set of bespoke simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe \texttt{stdpopsim}, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

@@

The degree to which modeling assumptions and using different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
% JK is this right? > 100 years doesn't sound very ancient to me.
suggested large ancient ($>100$ years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods (see
\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Furthermore, it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

One way to benchmark different methods for statistical inference in population
genetics is to apply the methods to simulated datasets where the true model
parameters are known. Methods that perform well will accurately infer these
parameters and will also provide reasonable estimates of uncertainty of the
estimated parameters. Indeed, simulations studies are a standard ingredient in
most papers proposing new methods in population genetics. However, these
simulation studies are often specifically devised to showcase the performance
of the novel method and may not adequately compare performance to competing
methods. Moreover researchers are generally forced to ``reinvent the wheel''
with respect to simulations, and are forced to reimplement previously
inferred scenarios independently. This can lead to errors being incorporated
during the implementation of the simulations. Lastly, the degree to
which different methods are compared can be quite variable.

One reason why simulation studies do not compare the performance of many methods
is that these simulation studies are challenging to carry out. There are several
practical impediments to carrying out the ideal benchmarking simulation studies.
First, while there are many published population genetic models available, their
parameters are often not easily accessible, usually presented in supplemental
tables. Translating these parameters into input for a given simulation engine can
be a time-consuming and error prone process. In addition, papers often present many
different models, and different authors may choose to select distinct models,
making comparison difficult. While improvements in simulation software and ever
faster computers have partially ameliorating this concern, generating the
simulated datasets can still be a time and computing resource bottleneck.

>>>>

\title{\stdpopsim: a community toolkit for population genetic simulation}

@@

\abstract{The explosion in population genomic data demands ever more complex
modes of analysis, and increasingly these analyses depend on sophisticated simulations.
Recent advances in population genetic simulation have made it possible to
simulate large and complex models, but describing such models for a particular
simulation engine remains a difficult and error-prone task.
Computational genomics researchers currently re-implement simulation models
independently for their own use, leading to duplication of effort and reductions in reproducibility.
Population genetics, as a field, also lacks rigorous standard benchmarks by which new tools for inference might
be measured. Here we describe a new resource, stdpopsim, that attempts to rectify this situation.
Stdpopsim is a community-driven, open source project, which provides a standard
catalog of published simulation models from a wide range of organisms,
and supports multiple simulation engine backends.
%KEL: I think we need another sentence to make it clearer on what exactly stdpopsim is...like what's available?
% PopSim assures correctness in our implementations through a mature quality control pipeline that is community-based.
Here we describe the stdpopsim resource to encourange an even broader
community of developers to participate and to demonstrate some examples of how
it can be used to systematically compare demographic inference methods.}
\date{\today}

@@

\section*{Introduction}
While population genetics has always used statistical methods to make inferences from data,
the degree of sophistication of the questions, models, data, and computational approaches
used have all increased over the past decade. Currently there exist myriad computational methods
that can infer the histories of populations, the distribution of fitness effects,
recombination rates, and the extent of positive selection in genome sequence data
(\cite{li2011inference,schiffels2014inferring,terhorst2017robust,eyre2009estimating,kim2017inference,
chan2012genome,lin2013fast,Adrion662247,alachiotis2012omegaplus,degiorgio2016sweepfinder2,
kern2018diplos,sugden2018localization}).
While en masse these methods have increased our understanding of the
impacts of genetic and evolutionary processes, very little has been done to systematically
benchmark the quality of inferences gleaned from computational population genetics.
This is in large part because ``ground-truth'' in population genetics comes from simulation,
not from empirical observations, and as a field there has been no agreement as to what should
constitute community standards or best practices. The general modus operandi to date has been to
introduce methods which are validated by a set of custom simulations produced by individual
groups. Beyond a lack of efficiency from duplicated effort, such a situation leads
to decreases in reproducibility and transparency across the entire field.
Here we describe \texttt{stdpopsim}, a community-driven catalog
of standardized simulations across a variety of common study organisms
that will empower the field and usher in the next era of genetics data analysis.

@@

The degree to which modeling assumptions and using different summaries of the
data can affect the ultimate results has been challenging to assess. Yet, there
have been several clear examples of different methods yielding fundamentally
different conclusions. For example, \MSMC methods applied to human genomes have
% JK is this right? > 100 years doesn't sound very ancient to me.
%KEL: Good catch--it should be >100,000 years ago!
suggested large ancient ($>100,000$ years ago) ancestral population sizes and
bottlenecks that have not been detected by SFS-based methods (see
\cite{beichman2017comparison}). These distinct models differ in how they fit various summaries of the
genetic variation data, suggesting that modeling choices can greatly affect the
performance of the inference. Furthermore, it is reasonable
to assume generally that specific methods will perform better than others under certain scenarios.
However, because these conditions are typically unknown, empirical researchers
lack principled guidelines for deciding which statistical method is best suited
to accurately answer their particular question. The need for empirical
guidance will only increase as researchers continue to generate more genomic
variation data from non-model taxa and wish to make inferences regarding
demography and selection.

One way to benchmark different methods for statistical inference in population
genetics is to apply the methods to simulated datasets where the true model
parameters are known. Methods that perform well will accurately infer these
parameters and will also provide reasonable estimates of uncertainty of the
estimated parameters. Indeed, simulation studies are a standard ingredient in
most papers proposing new methods in population genetics. However, these
simulation studies are often specifically devised to showcase the performance
of the novel method and may not adequately compare performance to competing
methods. Moreover researchers are generally forced to ``reinvent the wheel''
with respect to simulations, and are forced to reimplement previously
inferred scenarios independently. This can lead to errors being incorporated
during the implementation of the simulations. Lastly, the degree to
which different methods are compared can be quite variable.

One reason why simulation studies do not compare the performance of many methods
is that these simulation studies are challenging to carry out. There are several
practical impediments to carrying out the ideal benchmarking simulation studies.
First, while there are many published population genetic models available, their
parameters are often not easily accessible, usually presented in supplemental
tables, if at all. Translating these parameters into input for a given simulation engine can
be a time-consuming and error-prone process. In addition, papers often present many
different models, and different authors may choose to select distinct models,
making comparison difficult. While improvements in simulation software and ever
faster computers have partially ameliorated this concern, generating the
simulated datasets can still be a time and computing resource bottleneck.
